Sketch Recognition using Particle Swarm Algorithms

### 2009-07-17 trunk/smallpaper/SmallPaper.tex
changing the text for the  datasets

\documentclass{article}%{IEEEconf}[a4paper,10pt]
\usepackage{spconf}
\usepackage{amsmath}
%\usepackage[latin1]{inputenc}
%\usepackage[american]{babel}
%\usepackage[T1]{fontenc}
\usepackage[dvips,pdftex]{graphicx}
%\usepackage{multicol}
\usepackage{subfigure}
\usepackage{named}
%\usepackage{fancybox}
%\usepackage{natbib}

%\bibliographystyle{named}
\usepackage[dvips,pdftex]{hyperref}
%opening
\title{Sketch Recognition using Particle Swarm Algorithms}
%\title{A Study on Using Particle Swarm Algorithms in Symbol Recognition }
%\author{Maha El Meseery ;  Mahmoud Fakhr El Din ;  Samia Mashali ;  Magda Fayek and Nevin Darwish }
%\name{Maha El Meseery ;  Mahmoud Fakhr El Din ;  Samia Mashali ;  Magda Fayek and Nevin Darwish }

\twoauthors
  {Maha El Meseery, Mahmoud Fakhr El Din, Samia Mashali}
  %{}
	{Signals Processing Group \\ Computers and Systems Department \\
  Electronic Research Institute\\Cairo, Egypt\\
	melmeseery@eri.sci.eg, samia@eri.sci.eg}
  {Magda Fayek, Nevin Darwish}%\sthanks{The fourth author performed the work while at ...}
  %{}
	{Department of Computer Engineering\\
	Faculty of Engineering \\Cairo University\\Cairo, Egypt\\
	 magda@gmail.com,ndarwish@ieee.org}

%\address{ Electronic research inistiute}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Users draw strokes using a pen and the system should immediately interprets their strokes into objects that can be easily manipulated. This paper uses Particle Swarm Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. This paper uses Support Vector Machines(SVM) classifier which further improves the final recognition accuracy.  %u Final recognition has improved accuracy after the PSO algorithm using a SVM classifier.  
 %Hand drawn sketches are the natural method of exchanging thoughts between engineers and humans in general. Nevertheless, researchers barely give attention to improve current gesture and sketch recognition systems. In a sketch understanding system, users need to feel the freedom of paper and pencil along with the flexibility of computer manipulation. 
\end{abstract}
\begin{keywords}
Handwriting recognition, pattern recognition ,Document image processing
\end{keywords}
\section{Introduction}
%A picture speak a thousand words is also true for sketches. 
The field of symbol recognition has gained interest in the last few years. A wide variety of techniques were used either on segmentation process or symbol recognition. Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. %Engineers use paper and pencil design in early stages of design. 
Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of few advanced sketch recognition systems. %Sketch recognition system converts the drawings and informal graphs user draws into objects and symbols that can be used in further interaction with the system. 
Sketch recognition is divided into three main steps preprocessing, segmentation and symbol recognition. The preprocessing process saves user input stroke points from the hardware and collects basic information about the stroke then proceeds to remove noise and primary data calculation. In segmentation phase, the system divides the stroke into a set of simple geometrical primitives or segments. The third phase, sketch recognition, is clustering strokes and segments to formulate symbols that can be recognized by a classifier system. Finally, the sketch understanding phase uses priori information about the sketch to support identification of symbols by excluding any possible invalid symbols.

%the basic symbols are identified in consistent with the whole sketch. \\
%from the user information from the users converts the strokes the user draw into segments which is recognized into basic symbols that can be easily manipulated. Finally, the basic symbols are interpreted in consistent with the whole sketch. \\
This paper introduces a sketch recognition system. The presented system uses particle swarm algorithm (PSO) to optimally segment users strokes. Users can draw symbols using any number of strokes in any order they like. The system segments stroke user draws then cluster symbols together then passes them to the SVM symbol recognizer. The experiments in section \ref{sec:Experiments} shows that the proposed system can improve the final symbol recognition accuracy. 
%The remaining of this papers is organized as following; section \ref{Sec:RelatedWork} provides a brief literature review.  Section \ref{PSO} explains the general particle swarm algorithm which is used in the presented system. The proposed system block diagram along with a detailed system description is presented in Section \ref{Sysdisc}.  The experiments and results achieved are explained in Section \ref{sec:Experiments}. Finally, section \ref{ConclusionandFutureWork} states the conclusion and future work. 
%The next section will relate previous work in the field. The remaining of the paper will be as following; Section \ref{PSO} will explain the general particle swarm algorithm. Section \ref{Sysdisc} will describe the system block diagram and then section \ref{Prepross} will explain the preprocessing step, section \ref{seg} will describe the segmentation algorithm in details. Finally, section \ref{sec:Experiments} will describe the experiments that was preformed.\\
%\section{Literature Review}
%\label{Sec:RelatedWork}
%%%What is sketch recognitions. Why we need sketch recognition systems.

%To segment strokes
%In order to segment strokes, \cite{meanshift10} introduced a feature area for each primitive and then computed the segmentation error for different types of primitives based on this area.  %\citeauthor{HeloiseBeautification} used a dynamic programming algorithm to solve the segmentation problem \cite{HeloiseBeautification}. 
A hybrid algorithm was introduced in \cite{earlyprocess} where they generated different sets of segmentation based on both curvature and speed dominant points, this is followed by choosing a segmentation with the least error from generated hybrid set. Their system has a draw back as it only recognizes specific simple geometric shapes with a set of low level recognizers. Yu introduced a feature area for each primitive and then computed the segmentation error for different types of primitives based on the computed feature area. Their system achieved good accuracy in simple shapes (square, ellipse,...) but did not perform well in more complex shapes\cite{meanshift10}. Paulson and Hammond\cite{Paleosketch08} introduced a set of low level recognizers that were reported to achieve 98\%  accuracy but their system similar to all low level recognizers only identify a small set of simple shapes.

%Many different approaches have been investigated to achieve final recognition of symbols. Graph searching and template matching where used in \cite{GraphBased07} where a graph of the drawn symbol is generated and matched with a stored set of graphs templates to reach a classification decision. These systems may generate good performance but they are very computationally expensive. Geometrical and spatial descriptive language was used in \cite{SketchRead2007,Ladder30} where an architecture similar to compiler was used to recognize symbols. The drawback of this architecture is that the recognition system can only parse symbols that were hard coded using the descriptive language. New symbols are not recognized unless the description of the symbol is provided or generated using another system.  %%SVM and HMM classification was implemented in\cite{SVM300,HMM53} to correctly classify symbols.

Genetic algorithm was used by \cite{CruveDivisionSwarm} to optimally divide digital curves into lines and curves. Chen et al. uses digital curves scanned from paper as input to the system and did not take advantage of the curvature or local geometric properties of the digital curve. Yin \cite{PolygonApproximationPSO} used PSO to convert digital curves into polygons, our system adopts Yin's method and improves it by adding curvature and other local information while segmenting strokes to achieve better segmentations.% The next section presents the general particle swarm algorithm which is used in our proposed segmentation algorithms. % are presented in section \ref{subsubsec:Discreteparticleswarmalgorithm}
% Other techniques as  [ ] was used  in []. \\  
%In the next section the particle swarm algorithm will be explained. After that the system block description and detailed algorithms will be described. We will then proceed to describe the experiments preformed and the achieved results.  
%\subsection{Sketch Recognitions systems}
%Details of Sketch systems. 
%Related work and research in this area. 
%Writing on the pros and cons of current systems.  
%Write on the difficulties of the sketch recognition systems. 
  %Some tried to detect dominant or corner points based on data generated from speed   \cite{earlyprocess} and mostly used curvature of the stroke drawn \cite{earlyprocess,meanshift10}. 
%Particle swarm algorithms was used in \cite{} to fit a set of points into ellipse or higher order curves.
%The first set of features introduced to be used on gesture recognition was introduced by Rubine in \cite{gestureexample12}. later,  Spatial relations and features where used in  \cite{cognitivesketch18,SRGraph57}  . \citeauthor{GeometryAndDomain102} introduced ink density feature which they used.  The use of Zernike moments  was used  as the only features to the SVM classifiers in \cite{HeloiseBeautification}.  \cite{SketchRead2007} used a set of features varied from relative spatial features and some compositional features.  \\
%After segmentation researchers varies on the methods in which to continue.
%\section{Particle Swarm Algorithm}
%\label{PSO}
%%What is particle swarm algorithm and how it was used in related researches. 
%The main idea of Particle Swarm Algorithm (PSO) is to represent each agent with a particle from the solution space. Each agent moves the particle with a direction and velocity based on equation \ref{eq:Swarm}.
% %Equation [\ref{eq:Swarm}] shows how velocity and direction of each particle are computed
% \begin{equation}
%v_{ij}  = v_{ij}  + c_1 r_1 (lbest_{ij}  - p_{ij} ) + c_2 r_2 (gbest_{ij}  - p_{ij} )
%\label{eq:Swarm}
%\end{equation}
% where $r_1$ \& $r_2$ are random variables and $c_1$ \& $c_2$ are the swarm system variables.
%\begin{equation}
%%\[
%p_{ij}=p_{ij}+v_{ij},
%%\
%\end{equation}
%where $p_{ij}$ represent the $jth$ particle in the $ith$ agent and $v_{ij}$ is the velocity of the $jth$ particle in the $ith$ agent. After each iteration the global best $g_{best}$ particle and the agent local best $l_{best}$ particle are evaluated based on the maximum fitness functions of all particles. The solution is found after achieving a specific number of iteration or after an error threshold is achieved.
%The following equation \ref{eq:descrite}  
%\begin{equation}
%   P(i)\Leftarrow 
%\{
%\begin{array}{c} 
%1 \quad \quad if\quad r_{3}>t  \\
%
%0 \quad \quad if\quad r_{3}<t 
%\label{eq:descrite}
%\end{array}\}
%\end{equation}
% (where $t$ is a threshold and $r_{3}$ is a random variable), is used in order to change the general swarm algorithm into binary particle which handles particle values of either $0$ or $1$. 
%    
\section{System Details}
\label{Sysdisc}
%The stroke is the path of points from the instant the pen is down till it is up.
%Figure \ref{fig:Blockdiagram} 
The following sketch recognitions systems is divided into three main system steps are 1) preprocessing, 2) segmentation and 3) Recognition. The following sections provides a detail description of each step.
 %Due to sloppiness of users and hardware glitches the system preprocess the data before further processing. 
%The preprocessing phase is responsible of capturing the input data and removing the noise from it. The system then proceeds to estimate a set of possible dominant points to help in the segmentation process. % The time differences, direction, speed and curvature of all points of the stroke is computed along with some other geometrical data. The system then proceed to estimate a set of possible dominant points to help in segmentation process.  \\
  %then passes through the segmentation process which will try to 
 %The goal of the segmentation stage is to divide stroke into segments of either curves or lines.
%The segmentation is divided into two steps ellipse fitting and curve segmentation. %In the first step, the ellipse fitting process tries to fit the stroke into an ellipse. If the system fails, it passes the stroke to the second step; curve segmentation which consists of two PSO segmentation algorithms. The two algorithms generate two segmentations. The system then chooses the segmentation that has the minimum segmentation error.%generate the segmentations,  the minimum error will be the chosen segmentation. % is the ellipse fitting
 %the stroke first If the ellipse detection fails the stroke is passed to the segmentation algorithms which will pass it to the two PSO algorithms described below the segmentation with the minimum error will be the chosen segmentation.  The segmentation is then added to the set of un-recognized segments in the system. \\% This part is repeated for each stroke. 
 %The clustering algorithm starts to group segments together after the segmentation step. The system allows the user draws the symbol by using any number of strokes; a set of unrecognized segments is passed to the clustering algorithm to generate a symbol and compute a feature vector for it. A set of hybrid features of statistical, geometric and spatial features is used to generate the symbol feature vector. The final classification step, SVM classifier uses the feature vector computed to classify the segments into one of the previously trained classes.
  %  Or to determine the symbol of the given segments from the set of preciously trained symbols. 
 %with a symbol from the training set.% The system compute composite set of features some are statistical other are spatial features based on the type primitives. Finally, the strokes are classified into the corresponding classes.% the classifier identifies the strokes into a symbol from the set of known symbols.\\ 
%\begin{figure}[]
%%\begin{minipage}[b]{0.5\linewidth}
%	\centering
%		\includegraphics[scale=0.75]{images/block2.pdf}%Blockdiagram.pdf}
%	\caption{The system Block Diagram.} The figure shows the main system blocks. Preprocessing, segmentation, clustering, features extraction and classification. 
%	
%	\label{fig:Blockdiagram}
%%	\end{minipage}
%\end{figure}
\subsection{Preprocessing}
\label{Prepross}
%Some data is extracted from the points in the stroke.
 %It is noted that as the user draws a shape, the pen slows down near corners and picks up speed when drawing a straight lines. Therefore, the 
 Speed and time difference data were widely used in sketch understanding systems \cite{earlyprocess}. The direction and curvature data are used to determine the points with high angular changes along the path of points \cite{meanshift10}.  
 %speed is calculated as 
 In the presented system, we computed time difference, direction, speed and curvature of each point along the stroke. The speed is calculated as  $v=\Delta t/\Delta s$ where $t$ is the time difference between two points and $s$ is the length between them. The direction is calculated as the angle between two vectors and curvature is considered as the change in direction with respect to length i.e. $c= \Delta d/\Delta s$.
 %system compute speed, distance data for all points in the stroke. Curvature is computed using estimation used in [] where direction computed is angle between two lines. Curvature is 
 %$\Delta d/  \Delta S $ where d is the difference direction of point and s is difference in distance between points. 
 %It should be noted that, all these calculations are performed in real time while the user draws the strokes. The complexity of computation is $O(n)$ where $n$ is number of points.

 Figure \ref{fig:speed2} shows the computed speed data for the stroke drawn in Fig. \ref{fig:orignalStroke}. %If you look at the data curves you will see by looking at the data curves s
 It is noted that the dominant points are characterized by lower speed values and higher curvature and direction values(Fig. \ref{fig:speed2}). % the lower speed points correspond to vertex and dominant point's location. The higher the direction and curvature data correspond to location with higher change in curvature witch promote the location for vertices. \\
After the system computes all the speed, time difference and curvature data it proceeds to detect the points with low velocity and high curvatures. Using simple differentiations to detect local extreme points resulted in false points due to the non smooth curves. Hence, the system adopted a process presented by \cite{earlyprocess}, where the mean of the curve is calculated and used as threshold \textit{th}. The threshold \textit{th} is used to separate the curve into regions  $Region_i$; each region $Region_i$ is defined as the part between two intersection points of the threshold line and the curve. Regions higher than the threshold \textit{th} are processed to find the maximum point $Max(Region_i)$ of each region $Region_i$. The stroke points $p_i(x,y)$ that correspond to those maximum values are labeled as \textit{possible dominant points} $P_{pd}$. The system repeats this process to curvature, time difference and speed curves. All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. %They are then used as input to guide the next segmentation phase. 
%After the system computes all the speed, time difference and curvature data it proceeds to detect the points with low velocity and high curvatures. The system adopted the \textsl{average thresholding} process presented by \cite{earlyprocess}. This process is used to curvature, time difference and speed curves to generate a set of possible dominate points$P_{pd}$. These points are saved into a single array to be used as input to guide the next segmentation phase. %All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. They are then used as input to guide the next segmentation phase.%Using simple differentiations to detect local extreme points resulted in false points due to the non smooth curves. Hence, the system adopted the \textsl{average threshold} process presented by \cite{earlyprocess}. This process is used to curvature, time difference and speed curves. All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. They are then used as input to guide the next segmentation phase.  %, where the mean
%So instead the following procedure was used, firstly the mean of the curve is used as threshold for lower values then the high values are further investigated to find the extreme points. Hence, the mean 
%of the curve is calculated and used as threshold \textit{th}. The threshold \textit{th} is used to separate the curve into regions  $Region_i$; each region $Region_i$ is defined as the part between two intersection points of the threshold line and the curve. Regions higher than the threshold \textit{th} are processed to find the maximum point $Max(Region_i)$ of each region $Region_i$. The stroke points $p_i(x,y)$ that correspond to those maximum values are labeled as \textit{possible dominant points} $P_{pd}$.   \\
   %If we tried to differentiate the curve the result will be false threshold values we divide the data into regions of data higher or lower that the threshold. This will let us only look for high data only. The maximum of each region is then computed and reported as a possible vertex point. 
%\begin{figure}[]
%	\centering
%	
%		
%	\caption{Stroke Sample} An example of the stroke drawn in the system.
%
%\end{figure}
%The system repeats this process to curvature, time difference and speed curves. All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. They are then used as input to guide the next segmentation phase. 
\begin{figure}[]
%\begin{minipage}[b]{0.9\linewidth}
	\centering
		\subfigure[ The Speed of data] {	\label{fig:speed2}\includegraphics[scale=0.2]{images/speed22.PNG}}	
			\subfigure[Stroke Sample] {		\label{fig:orignalStroke} \includegraphics[scale=0.3]{images/orignalStroke.pdf}}
	%	\subfigure[ Direction ] {\includegraphics[scale=0.15]{images/direction2.pdf}}
					%\end{minipage}
		%			\hspace{20 mm}
					%\begin{minipage}[b]{0.5\linewidth}
		%	\subfigure[The Curvature] {\includegraphics[scale=0.29]{images/curvature2.pdf}}
	%		\subfigure[ Time Difference] {\includegraphics[scale=0.15]{images/time2.pdf}}
%			\end{minipage}
	\caption{Data Curves} An example of the stroke drawn in the system with the curve that shows the speed of the points in the stroke.% in fig. \ref{fig:orignalStroke}.% b) Shows the curvature curve for the same stroke.  %{The data of the stroke}
\end{figure}
%Description of input data. 
%how to calculate speed, curvature, area, bounding box.
%how to remove noise. 
%Finally how to compute primarily dominant points. 
\subsection{Segmentation}
\label{seg}
After computing the primary data the system tries to segment the stroke into a set of primitives. The segmentation algorithm first attempts to fit the stroke points into a curve or an ellipse uisng a minimum square error fitting algorithm. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed into two particle swarm algorithms that divides the stroke to either lines or lines and curves. The algorithms take the stroke points along with the possible dominant points $P_{pd}$ computed then produce a set of dominant points which are connected with either lines or curves. The next section describes the ellipse detection algorithm and the two particle swarm algorithms used to divide the stroke.\\ %The next section describes the two particle swarm algorithms used to divide the stroke. %(see Fig. \ref{fig:Blockdiagram})

>>>>

\documentclass{article}
\usepackage{spconf}
\usepackage{amsmath}
\usepackage[dvips,pdftex]{graphicx}
\usepackage{subfigure}
\usepackage{named}
\usepackage[dvips,pdftex]{hyperref}
\title{Sketch Recognition using Particle Swarm Algorithms}
\twoauthors
  {Maha El Meseery, Mahmoud Fakhr El Din\sthanks{Associate Professor}, Samia Mashali\sthanks{Professor}}
	{Signals Processing Group \\ Computers and Systems Department \\
  Electronic Research Institute\\Cairo, Egypt\\
	melmeseery@eri.sci.eg, samia@eri.sci.eg}
  {Magda Fayek\sthanks{Associate Professor}, Nevin Darwish\sthanks{Professor}}
	{ Computer Engineering Department\\
	Faculty of Engineering \\Cairo University\\Cairo, Egypt\\
	magdafayek@gmail.com,ndarwish@ieee.org}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Users draw strokes using a pen and the system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. This paper uses Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.  
\end{abstract}
\section{Introduction}
The field of symbol recognition has gained interest in the last few years. Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of a few advanced sketch recognition systems.

Sketch recognition is divided into three main steps preprocessing, segmentation and symbol recognition. The preprocessing phase captures user input stroke points and collects basic information about the stroke then proceeds to remove noise and compute basic statistical and geometrical information. In segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier system.

A wide variety of techniques were suggested for sketch recognition.  A hybrid algorithm was introduced in \cite{earlyprocess} where they generated different sets of segments based on both curvature and speed dominant points, followed by choosing a segmentation with the least error from a generated hybrid set. Their system is however limited to recognizing only specific simple geometric shapes with a set of low level recognizers. Each low level recognizer is designed to recognize only one geometric shape using spatial and geometric information extracted from the stroke.  Yu \cite{meanshift10}introduced a feature area for each primitive and then computed the segmentation error for different types of primitives based on the computed feature area. His system achieved good accuracy in simple shapes (square, ellipse,...) but did not perform well in more complex shapes. Paulson and Hammond\cite{Paleosketch08} introduced a set of low level recognizers that were reported to achieve 98\%  accuracy but their system, similar to all low level recognizers, only identify a small set of simple shapes. A genetic algorithm was used by \cite{CruveDivisionSwarm} to optimally divide digital curves into lines and curves. Chen et al.\cite{CruveDivisionSwarm} uses digital curves scanned from paper as input to the system and did not take advantage of the curvature or local geometric properties of the digital curve.

This paper introduces a new sketch recognition system, which uses particle swarm algorithm (PSO) to optimally segment user's strokes. Users can draw symbols using any number of strokes in any order they like. The system segments each stroke using two different PSO algorithms. Segmentation is based on dividing the digital curve into polygon\cite{PolygonApproximationPSO}. The algorithm enhances its performance by adding curvature and other local information while segmenting strokes to achieve better results. To handle complex strokes, our system uses a second segmentation PSO algorithm to segment strokes into a set of lines and circular arcs. After segmentation, a symbol features vector composed of different statistical, geometrical and spatial features is deduced. The final classification step, SVM classifier uses the computed feature vector to classify the segments into one of the previously trained classes.

 The remaining of the paper is as following; Section \ref{Sysdisc} describes the system in details.  Sections \ref{Prepross}, \ref{seg} and \ref{sec:Recognition} provide details on preprocessing, segmentation and recognition process respectively. Section \ref{sec:Experiments} describes the experiments that were preformed. We conclude in Section \ref{ConclusionandFutureWork}.
\section{The proposed system}
\label{Sysdisc}
 The suggested sketch recognitions system is divided into three main steps 1) preprocessing, 2) segmentation and 3) Recognition. The following section provides a detailed description of each step.
\subsection{Preprocessing}
\label{Prepross}
 Speed and time difference information were widely used in sketch understanding systems \cite{earlyprocess}. It was noted that time difference provides more distinct maxima than the speed information. Agar \cite{polygonfeedback31} mentions that the pointing device (for example the mouse) sampling rate is the reason for this phenomena. 
 In the presented system, we computed time difference, direction, speed and curvature of each point along a stroke. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ is calculated as the angle between two vectors and curvature is considered as the change in direction with respect to length i.e. $c= \Delta d/\Delta s$.  Dominant points are characterized by lower speed values and higher curvature and direction values.

After the system computes all the speed, time difference and curvature information it proceeds to detect the points with low velocity and high curvatures. Using simple differentiation to detect local extreme points resulted in false points due to the non smooth curves. Hence, the system adopted a process presented by \cite{earlyprocess}, where the mean of the curve is calculated. Then a threshold \textit{th} is used to separate the curve into regions $Region_i$; each region $Region_i$ is defined as a range of points, where the curve values are either above or below the threshold \textit{th}. Those regions are further processed to find the maximum point $Max(Region_i)$ of each region $Region_i$. The stroke points $p_i(x,y)$ that correspond to those maximum values are labeled as \textit{possible dominant points} $P_{pd}$. The system repeats this process for curvature, time difference and speed information. All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. Figure \ref{fig:LabelsPPD} shows the particles labeled as Possible dominate point $P_{pd}$ by the preprocessing, it is noticed the redundancy of some $P_{pd}$ points. % (as shown are redundant)
\begin{figure}
	\centering
		\subfigure[Posible domninate point]{ 	\label{fig:LabelsPPD}	\includegraphics[scale=0.5]{images/ppd.jpg}}
		\hfill
			\subfigure[DPSO algorithm encoding]{	\label{fig:LabelsPSO} \includegraphics[scale=0.5]{images/pso1.jpg}}			
	\caption{Segmentation}% a) Possible dominate point b) Particle encoding  
	\label{fig:pso1}
\end{figure}
\subsection{Segmentation}
\label{seg}
In the segmentation stage the stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{chernov-2003}. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two particle swarm algorithms that divide the stroke to either lines or lines and curves. The algorithms takes the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing then produce a set of dominant points which are connected by either lines or curves. The next section describes the ellipse detection algorithm and the two particle swarm algorithms used to divide the stroke.
 \begin{figure}
	\centering
		\includegraphics[scale=0.55]{images/segblock.jpg}
	\caption{Segmentation Block}
	\label{fig:segblock}
\end{figure}
\subsubsection{Ellipse Detection} 
The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov-2003}.  
\subsubsection{Discrete particle swarm algorithm (\textit{DPSO})}
\label{subsubsec:Discreteparticleswarmalgorithm}
Two DPSO algorithms are used to generate segmentations for each stroke the user draws. For each generated segmentation evaluated an error is determined. The segmentation with the minimum error value is chosen as the best stroke segmentation (Figure \ref{fig:segblock}). The problem definition is the same in both algorithms but they differ in the method they use to compute fitness and error functions.

\textsl{Problem definition:} The input stroke with $N$ points can be represented by set $S = \left\{ {x_1 ,x_2  \ldots x_N }\right\}$ where $x_i$ is the location of the point $i$. The swarm algorithms consist of $M$ agents which are represented by the set  $A = \left\{ {P_i \left| {i = 1,2 \cdots M} \right.} \right\}$ where $P_i$ is a single solution particle from the solution space. Each particle decodes the problem using a binary array with the same length $N$ as the input stroke.

Therefore, the system represents each particle $P_i$ by $P_i = \left\{ {p_{ij} \left| {j = 1,2 \cdots N} \right.} \right\}$ where $p_{ij}$ has only two values 0 or 1. $p_{ij}=1$ means that point $j$ is a dominant point. Figure\ref{fig:LabelsPSO} shows a particle encoding in the DPSO system.

\textsl{Fitness function:} The fitness function and error calculation are different in each of the two \textit{DPSO} algorithms.

For the \textit{first algorithm \textsl{AlgS1}}, 
 the arc $\widehat{x_ix_j}$ is defined as the consecutive set of points from point$x_i$ to point $x_{j}$ as in $x_i,x_{i+1} \cdots,x_j$. The line $\overline{x_i x_j} $ as the straight line connecting point $x_i$ to point $x_j$. The approximation error is computed by the equation \ref{eq:ErrorSwarm1}. The graphical meaning of the error is shown in Figure\ref{fig:LabelsPSO}.
\begin{equation}
E=\sum\nolimits_{i = 0}^M e ( \widehat{x_ix_{i+1}},\overline{x_i x_{i+1}})
\label{eq:ErrorSwarm1}
\end{equation} where $M$ is the number of dominant points in this solution as generated by the swarm algorithm. The error $e ( \widehat{x_ix_j},\overline{x_i x_j})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$. The fitness is computed using equation \ref{eq:fitnessSwarm1}\begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{20}c}
   { - E/\varepsilon N} & {ifE > \varepsilon ,}  \\
   {D/\sum\limits_{j = 1}^N {p_{ij} } } & {otherwise}  \\
\end{array}} \right.
\label{eq:fitnessSwarm1}
\end{equation} where $N$ is the number of points in the stroke, $D$ is the number of points in the solution that was previously labeled as a possible dominant point ($P_{pd}$), $E$ is the computed error and $\varepsilon$ is the error threshold. It should be noticed that when the error is larger than the threshold $\varepsilon$ the fitness is given a -ve value to lower the fitness value of the solution. Otherwise the system favors the lower number of vertices.

The \textit{second algorithm \textsl{AlgS2}} has the same problem formulation but different fitness and error functions are used. An attempt is made to fit each segment into line or circular arc, the error of both circle and line fit estimations are computed for each segment $S_i=\widehat{x_ix_j}$, the approximation with the lower error value is the chosen approximation of this segment $S_i$\cite{CruveDivisionSwarm}. The sum of the approximation error of all segments is the total error of the particle.  The total error of the particle is computed by equation \ref{eq:errorSwarm2}
 \begin{equation}
E=\sum\nolimits_{i = 0}^M e(D_i) 
\label{eq:errorSwarm2}
\end{equation}where $M$ is the number of segments in the solution as generated by the swarm algorithm, $D_i$ is the minimum approximation error of curve and line approximations $min(d_c,d_l)$ as computed by \cite{CruveDivisionSwarm}.  The fitness is computed by the equation \ref{eq:fitnessSwarm2} \begin{equation}
\max fitness(P_i ) = \frac{1}{{E \times M^k }}
\label{eq:fitnessSwarm2}
\end{equation} where $E$ is the error and $M$ is the number of segments and $k$ is a parameter tweaked to get minimum number of segments. The larger k is, the more effect the number of segments will have. For our system, k is selected to be 0.5\cite{CruveDivisionSwarm}.

After each loop of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedure. For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominate points $P_{pd}$. After that the particles are tested to make sure that the distance between every two successive dominate point is larger than the constant $min_D$. If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The features used consist of Rubine feature set,  Zernike moments of order 21 \cite{HeloiseBeautification}, ink density as well as some structural and spatial information like number of perpendiculars lines ,number of parallel lines and number of different types of primitives in each symbol. After computing the features the symbol is introduced to the (SVM) classifier. 
\section{Experimental}
\label{sec:Experiments}

The dataset collected by Hse and Newton\cite{HeloiseBeautification} is used to test the presented system.  Figure \ref{fig:symbolSet} shows a set of the shapes used in the data set. The datasets were divided into training and test sets. Five different splits for the training and test data are generated from each dataset. The results displayed are the average recognition accuracy of the five splits. The accuracy is computed as the number of correctly recognized samples divided by the total number of samples in the test.

 \begin{figure*}[]
	\centering
	\begin{minipage}[b]{5cm}	
\begin{center}
\subfigure [The Symbol Set] {\label{fig:symbolSet}
\fbox{ \parbox{2cm}{% 
		\includegraphics[scale=0.2]{images/symbolSet.PNG}	}}}

		\subfigure[Algorithm comparison]{\includegraphics[width=5cm]{images/test1smallpaper.jpg} 	\label{fig:test1}}
		\end{center}
		\end{minipage}
		\hfill
  \subfigure [Symbols Comparison] {\includegraphics[width=9cm]{images/test2smallpaper.jpg} \label{fig:test2}}
	\caption{Experiments results :} a) the figure shows a set of the symbols used in the dataset.   b)The results of comparing the algorithms. The graph displays the recognition rate of symbols. c) The graph shows the recognition rate of each symbol using different algorithms. 
\end{figure*} 
We performed two experiments to test the system. Firstly we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). Figure \ref{fig:test1} shows the accuracy achieved by each algorithm. The two swarm algorithms were tested with and without the ellipse fitting module. Including the ellipse detection module improves the results with both the \textit{DPSO} algorithms.% The results show that both \textit{DPSO} algorithms achieve better result than other algorithms.

The second experiment we implemented was to test the effect of symbol complexity and type on the recognition rate. Figure \ref{fig:test2} shows the accuracy of each symbol, it is clearly noted that symbols that have only line segments achieve higher accuracy rate than other symbols. Figure \ref{fig:test2} also indicates that algorithm \textsl{AlgS1} alone achieve better performance than algorithm   \textsl{AlgS2} in the symbols that consist only of lines. This is natural as algorithm \textsl{AlgS1} divides strokes into line segments only but \textsl{AlgS2} is able to divide strokes into lines and curves based on the minimum error of the segment itself.  
 Combining both algorithm \textsl{AlgS1} and \textsl{AlgS2} improved the recognition rate of all symbols. The penalty for this increased performance is in the computational time needed to run both swarm algorithms. 
\section{Conclusion and Future Work}
\label{ConclusionandFutureWork}
This paper presented a new approach to sketch recognition using PSO. The system uses both speed and curvature information which help improving the \textit{DPSO}. It is noted that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The tradeoff between accuracy achieved and time complexity must be further investigated to achieve better results. The use of statistical moments and structural features improves the recognition rate. The system was tested on 13 different symbols and achieved an accuracy of 91.5\%. The system does not depend on low level recognizer but rather on a set of high level features. This makes the system easily expandable to more and more symbols.

 A possible extension of this research is to complete the clustering algorithm for fully automated sketch recognition. Currently, the system process one symbol at a time, an enchantment would be to allow users to draw more than one symbol at a time and incorporate a method for separating symbols. Other area of enhancements is the features extraction methods. Introducing more spatial and geometrical features is believed to improve classifications.  
\bibliographystyle{IEEEbib}
\bibliography{../../neededfiles/Bibliographies/Mybibliography}
\end{document}
